Run AwAI: StarCraftII Adversary Avoidance Using Neural Networks and Genetic Algorithms

### 2018-06-04 paper.tex
update paper, add figures

%% TODO: Why our approach vs other approaches
%% TODO: Random action is a fairly robust approach to the problem space
%% TODO: Include convergence Graphs in the results section
%% TODO: Talk about the mutation rates in the genetic algorithm

@@

\subsection{Simulations}
We run the neural networks through a simulation to compute their fitness. The
simulations run through the PySCII interface, which runs our agent in a
simulation against pre-built AI.

@@

\begin{figure}[H]
\centering
%\includegraphics[width=0.45\textwidth]{chart}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Control and data flow through the Run AwAI controller.}
\end{changemargin}
\end{figure}

@@

\subsection{Neural Network}
We started by using an explicit neural network implementation, which we chose
because we were curious if the flexible topologies would lead to a more
efficient neural network. We were inspired by the NEAT algorithm , which has
shown promise in other video game domains. However we choose a simpler
topologically explicit implementation to simplify the development of our
algorithm. We ultimately settled on an acyclic graph representation, which
allowed us the freedom to add nodes and allowed simple manipulation of
connection weights.

@@

To get convergence, we spent a lot of time looking at fitness functions. We
considered many different
possibilities, but finally settled on a measure of the number of steps that our
agent took during the simulation. We figured that this would be a good measure
of our agent's ability to stay alive, and we thought that time alive was a good
indicator of the agent's ability to accomplish the goal of running away. We
computed the fitness $F$ using the following formula:
\begin{align}
  F(\bm{i}) = \mu - \eta\sqrt{\frac{1}{|\bm{\mathcal{E}}|}
  \sum_{e \in \bm{\mathcal{E}}}\left(\mathcal{S}(e) - \mu\right)^2} 
\end{align}
where
\begin{align}
  \mu = \frac{1}{|\bm{\mathcal{E}}|} \sum_{e \in \bm{\mathcal{E}}} \mathcal{S}(e)
\end{align}
and $\bm{\mathcal{E}}$ is the set of all episodes run by the PySCII simulation,
$\mathcal{S}(e)$ is the number of steps taken by the agent in the episode, and
$\bm{i}$ is a an individual of the population. $\eta$ is a hyper-parameter we
use
to fine tune the convergence of our network. We are effectively calculating the
average number of steps and penalizing the individuals based on the standard
deviation from the mean of the episodes. This serves to mitigate the effects
presented by individuals that get ``lucky'' and have a good trial runs that
offsets their average.

@@

\subsection{PySC2 Interface}

@@

\section{Results}
%% TODO: Talk about choice of hyper paramters, population size, episode count
%% mutation constants, eta in fitness function

\section{Conclusion}
Different selection types?

\subsection{Improvements}

\subsection{Future Work}
This project can certainly be extended in many ways, for example, we could
extend the output vectors to include offinsive manuvering by our agent, and
train the agent to be agressive by considering the numebr of adversaries killed
by our agent during the simulation.

>>>>

%% TODO: Why our approach vs other approaches
%% TODO: Random action is a fairly robust approach to the problem space
%% TODO: Include convergence Graphs in the results section
%% TODO: Talk about the mutation rates in the genetic algorithm
%% TODO: Reference the fact that neuro evolution is good in high dimensional
%%    and continuous state spaces (ref neat paper)

@@

\subsection{Simulations}
To compute the fitness for each neural network in the population, we run the
network through a simulation. The simulation is run in StarCraftII, and our
simulation controller is built on top of the PySCII interface, and simulates
a ``game'' with our agent and a group of adversaries, which are controlled by
the ingame AI that come with the game.

Each simulation is called an episode, and we ran the networks through 5-10
episodes each so that we could get a representative sample of the amount of time
that each nerual network was able to survive.

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{chart}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Control and data flow through the Run AwAI controller.}
\end{changemargin}
\end{figure}

@@

\subsection{Neural Network}
We started by using an explicit neural network implementation, which we chose
because we were curious if the flexible topologies would lead to a more
efficient neural network. We were inspired by the NEAT algorithm \cite{NEAT}, which has
shown promise in other video game domains. However we choose a simpler
topologically explicit implementation to simplify the development of our
algorithm. We ultimately settled on an acyclic graph representation, which
allowed us the freedom to add nodes and allowed simple manipulation of
connection weights.

@@

To get convergence, we spent a lot of time looking at fitness functions. We
considered many different
possibilities, but finally settled on a measure of the number of steps that our
agent took during the simulation. We figured that this would be a good measure
of our agent's ability to stay alive, and we thought that time alive was a good
indicator of the agent's ability to accomplish the goal of running away. We
computed the fitness $F$ using the following formula:
% \begin{align}
%   F(\bm{i}) = \mu - \eta\sqrt{\frac{1}{|\bm{\mathcal{E}}|}
%   \sum_{e \in \bm{\mathcal{E}}}\left(\mathcal{S}(e) - \mu\right)^2} 
% \end{align}
% where
% \begin{align}
%   \mu = \frac{1}{|\bm{\mathcal{E}}|} \sum_{e \in \bm{\mathcal{E}}} \mathcal{S}(e)
    %   \end{align}
\begin{align}\label{eq:fitness}
  F(\bm{i}) = \frac{1}{|\bm{\mathcal{E}}|} \sum_{e \in \bm{\mathcal{E}}} \mathcal{S}(e) - \eta\sqrt{\frac{1}{|\bm{\mathcal{E}}|}
   \sum_{e \in \bm{\mathcal{E}}}\left(\mathcal{S}(e) - \frac{1}{|\bm{\mathcal{E}}|} \sum_{e \in \bm{\mathcal{E}}} \mathcal{S}(e)\right)^2} 
\end{align}
and $\bm{\mathcal{E}}$ is the set of all episodes run by the PySCII simulation,
$\mathcal{S}(e)$ is the number of steps taken by the agent in the episode, and
$\bm{i}$ is a an individual of the population. $\eta$ is a hyper-parameter we
use
to fine tune the convergence of our network. We are effectively calculating the
average number of steps and penalizing the individuals based on the standard
deviation from the mean of the episodes. This serves to mitigate the effects
presented by individuals that get ``lucky'' and have a good trial runs that
offsets their average.

@@

\subsection{PySC2 Interface}
Our simulation learning environment was based on an existing environment
included in the PySC2 library\cite{pysc2}. We use the ``Defeat Roaches'' mini
game as the baseline for our learning. Once we had the map, we used the
StarCraftII map editor to tweak the mini game to better suit our needs. We
created three maps that were used in testing. The defualt map is a simple square
map which spawns a group of adversarialy ``roachs'', and a group of ``marines'',
who are controlled by our agent. The marines and roaches are spawned in the
same place every time, which gives us a consistent platform to test our agent.

The second variation is similar to the first, except that there is only one
``roach'' and one ``marine''. This map type highlights the capabilities of our
agent in a more intimate setting. The group of marines has the advantage of
being able to get ``lucky'' and have some members survive, but the one on one
variation forces the neural network to work well from the beginning.

@@

\section{Results}
We ran several simulations with various choices of hyperparameters, and our
results are promising. We saw measurable improvement in preformence after just
a few iterations of the algorithm, which gave us a steep initial learning curve
before the improvements started to taper off.

\begin{figure}[H]\label{fig:fitness graph}
\centering
\includegraphics[width=0.45\textwidth]{fig_1}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Graph of average fitness (blue) and top fitness (orange) for each
    generation in Trial 1. Top fitness is the fitness of the most fit individual
    at the end of each generation's simulations. Average fitness is calculated
    by averaging \eqref{eq:fitness} over all individuals $\bm{i}$ in the
    population. The breeding constant is set to 15, node generation rate is set
    to 0.3, new connections at 0.2 and non structural mutations at a rate of
    0.4}
\end{changemargin}
\end{figure}

We see in Figure \ref{fig:fitness graph} an artifact from our choice of genetic
algorithm. Because we keep the top individual from each generation in the gene
pool, we see that the top individual has palteaued early, and no new individual
is achieving a higer fitness. We see that the average is countinually decreasing
though, which will increase the chances that an individual will be bred that has
a higher fitness than the individual that is dominating the maximum fitness.

We can also see that the average sometimes plateaus as well. This is due to a
homonginization of the gene pool, where the children individuals are very
similar to their parents. This is a case of finding a local maximum. Since we
favored an exploitive algorithm, this is to be expected, however we see that the
population can still achieve higher fitness meausures which means that the
algorithm still retains some amount of exploratory behavior.

\begin{figure}[H]\label{fig:fitness graph}
\centering
\includegraphics[width=0.45\textwidth]{fig_2}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Graph of average fitness (blue) and top fitness (orange) for each
    generation in Trial 1. Top fitness is the fitness of the most fit individual
    at the end of each generation's simulations. Average fitness is calculated
    by averaging \eqref{eq:fitness} over all individuals $\bm{i}$ in the
    population. The breeding constant is set to 15, node generation rate is set
    to 0.5, new connections at 0.5 and non structural mutations at a rate of
    0.75}
\end{changemargin}
\end{figure}

In our second trial, we see that the average fitness increases much more
regularly. However, despite having more generations, trial two did not reach the
same average fitness as trial one did. One possible reason for this is since our
mutation rates are higher, our algorithm favored a more exploratory approach,
which lead to a more consistant incresase in fitness, but did not reach as high
of an average fitness.

%% TODO: Talk about choice of hyper paramters, population size, episode count
%% mutation constants, eta in fitness function

\section{Conclusion}
Different selection types?
\blindtext

\subsection{Improvements}
\blindtext
\blindtext

\subsection{Future Work}
This project can certainly be extended in many ways, for example, we could
extend the output vectors to include offinsive manuvering by our agent, and
train the agent to be agressive by considering the number of adversaries killed
by our agent during the simulation.

Another possibility that we did not explore, was the effect of removing neurons
from the network. Neural networks can often be plauged by overfitting input
data, and by removing neurons we hypothesize that any overfitting we encounter
could be mitigated. We would remove these neurons in the same way that we add
neurons, that is, non-deterministically as a possible mutation. In the study of
biological creatures, we often see either total dropping of traits from the
population genome or the transition to vestigial genetic structure
\cite{lehman}. In keeping with this, instead of outright removing the neuron, we
could significantly decrease the strength of the connections to it, essentially
removing it from the network. If the neuron in question was indeed problematic,
we would see it countinue to evolve very low weights, essentially pruning the
neuron and rendering it vestigial.

We could also track the lifespan of certain individuals and kill them off after
a certain number of generations. We found that the algorithm would sometimes
plateau with a very fit individual for a period of time, and it would be
interesting to see what effects a notion of ``lifetime'' would have on the
algorithm as a whole.
